Early environment and numerical discrimination in a lizard

### 2024-10-09 ms.qmd
Updates

```{r, packages}
#| label: packages
pacman::p_load(tidyverse, flextable, emmeans, DHARMa, brms, here, ggplot2, lme4, zoo, lmerTest, broom, tidybayes, forcats, MASS,
fitdistrplus, ggh4x, PupillometryR, cowplot, png, grid, remotes, ggthemes)
```

@@

## Abstract
From social behaviour to navigating their environment, quantitative abilities can be crucial in animals' life. Early life conditions are predicted to impact brain development, and quantitative competence may not be exent of these effects. Early thermal environment or glucocorticoids (GCs) - stress-related hormones - transmitted by the parents can play a prominent role in shaping cognition through effects on brain physiology. Furthermore, temperature can lead to increased levels of GCs that are passed on to the offspring, emphasizing the need to study the combined effects of these two factors. Here, we investigated the effects of prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature on numerical discrimination in the common garden skink (_Lampropholis guichenoti_). We manipulated CORT levels and temperature during incubation and then assessed numerical abilities through a spontaneous choice test with food as the relevant stimulus. Employing a repeated measures design, we subjected lizards to five numerical tests each differing in the total number of items and the ratios between the two choices (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 2 VS 4). Contrary to our expectations, we found no evidence of numerical discrimination or any effect of treatments on lizards' behaviour. These findings suggest that _L. guichenoti_ may not use numerical cues to make foraging decisions and that early conditions do not affect their behaviour towards different quantities of food.

## Significance statement
Discriminating between quantities is crucial for animal behaviour, yet the impact of early environmental factors on the development of quantitative competence remains largely unexplored. In this study, we investigated for the first time the effects of prenatal corticosterone exposure and incubation temperature on numerical discrimination in a reptile. Contrary to our predictions, we did not observe any evidence of numerical discrimination regardless of incubation conditions. Our results align with previous studies showing that reptiles consuming live prey struggle to differentiate prey numbers in spontaneous choice tests. These results open new avenues for exploring the evolution of numerical abilities in reptiles and identifying factors influencing the development of quantitative skills in this group. Further research should explore alternative methods, such as trained tasks, to better understand the cognitive mechanisms underlying quantitative competence in reptiles.

## Introduction  
The ability to discern between quantities is a fundamental cognitive skill widely observed across the animal kingdom [@stancher2015numerical; @vonk2012bears; @beran2016capuchin]. Numerical cognition is adaptive in various ecological contexts, contributing to social decisions, mate selection, or foraging [@lucon2017individual; @nieder2018evolution]. For instance, female lions (_Panthera leo_) use numerical information to assess the risk of confronting rival groups [@mccomb1994roaring], male adult mealworms (_Tenebrio molitor_) select sites with scents of more females [@carazo2009quantity], and red-backed salamanders (_Plethodon cinereus_) prefer tubes containing larger numbers of prey [@uller2003salamanders]. Given the adaptive advantages of numerical and quantity discrimination, research has extensively explored the ability of different species to differentiate between quantities under different contexts and paradigms.

Numerical competence is often evaluated by considering the types of cues animals use [@plotnik2019elephants; @cooper2024tell], the ecological contexts in which these abilities are deployed [@mccomb1994roaring; @chittka1995can; @uller2003salamanders; @carazo2009quantity], and the neural mechanisms that underlie them [@nieder2005counting]. The latter have demonstrated that non-verbal numerical representations in mammals activate an extensive cortical network, with the prefrontal cortex and the intraparietal sulcus being particularly relevant [@nieder2005counting]. This connection between numerosity and brain physiology underscores the importance of understanding how brain development influences numerical abilities.

Animals' brains are particularly sensitive to environmental inputs during early stages of development [@coomber_independent_1997; @zhu_prenatal_2004], with long-lasting effects on cognitive abilities [@zhu_prenatal_2004; @amiel_egg_2014]. For instance, glucocorticoids (GCs), the main groups of hormones mediating the physiological stress response [@sapolsky_how_2000], can alter animals' performance in spatial, associative, or reversal learning tasks if elevated during early stages of development [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. Prenatal exposure to GCs is typically associated with impaired learning skills, but these effects have been shown to vary depending on factors such as dose, subjects' sex, or type of cognitive test employed [@szuran_water_1994; @farrell_developmental_2015-learn; @bebus_associative_2016; @haq_prenatal_2021]. Nevertheless, little is known about how exposure to GCs during development can affect quantitative abilities.

At the same time, the early thermal environment plays a prominent role in brain development and cognitive abilities in ectotherms [@coomber_independent_1997;@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_incubation_2018; @vila_pouca_quantity_2019]. For example, Port-Jackson sharks (_Heterodontus portusjacksoni_) incubated at elevated temperatures learned faster in a quantity discrimination task [@vila_pouca_quantity_2019]. However, the effects of early thermal environment on numerical abilities in other taxa remains largely unexplored.

Given that abrupt temperature changes can increase GC levels in vertebrates, leading to elevated GCs transmitted to the offspring [@racic_effects_2020; @Crino_2023], it is essential to study the combined effects of early-life GCs and thermal conditions on cognitive development. Understanding how these factors influence quantitative abilities could reveal important insights into the ecological relevance of numerosity. To fill this gap, we used the common garden skink (_Lampropholis guichenoti_) as a model to test the effects of elevated prenatal corticosterone (CORT) - the main GC in reptiles [@Crino_2023] - and incubation temperature on numerical discrimination abilities.

Traditionally, research on numerical abilities has been mainly focused on mammals, birds, and fishes [@nieder2005counting; @agrillo2014spontaneous], but recent studies demonstrate that reptiles are also competent in numberical abilities. For example, red-footed turtles (_Chelonoidis carbonaria_) rapidly learned to associate colour with higher volumes of food [@soldati2017long], Hermann’s tortoises (_Testudo hermanni_) proved to discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous], and freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) rapidly learned a quantity discrimination task [@lin2021superior; @lin2024trained]. The italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. In contrast, Iberian rock lizards (_Iberolacerta cyreni_) preferred larger quantities of crickets when presented five versus ten items [@recio_prey_2021]. Indeed, the literature shows evidence for basic quantitative abilities in reptiles that may vary with the type of cue and the test employed.

Reptiles' cognition is significantly influenced by the environment in which they develop. For example, incubating lizards at elevated temperatures improves cognitive abilities consistent with elevated neuronal density in some areas of the brain [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017], but impairs cognition at temperatures above lizards’ natural incubation thermal range [@dayananda_incubation_2017; @abayarathna_effects_2020]. The effects of prenatal CORT on cognitive abilities in reptiles is less understood, but in other taxa, high levels of GCs are typically related to cognitive impairments [@szuran_water_1994; @zhu_prenatal_2004]. However, the combined influence of prenatal CORT and thermal environment on cognition remains unknown, particularly on numerical abilities.

Our study aimed to address this question by manipulating CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubating them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. After hatching, we assessed the lizards' numerical discrimination abilities in a spontaneous choice test, where they were presented with different numbers of crickets. Based on other studies that explored the cognitive effects of prenatal GCs or thermal early environment in reptiles and other taxa, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. Additionally, we predicted that increased CORT levels would impair numerical competence [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures. In this sense, a common observation in nature is that the ability to detect a change in stimulus intensity depends on the relative, not absolute, difference between the stimuli ['Weber's law'; @agrillo2014spontaneous]. Consistent with Weber's law, we predicted that lizards incubated at higher temperatures would successfully discriminate between quantities even at high ratios (e.g. 3 VS 4), whereas even the lowest ratios (e.g. 1 VS 4) could be challenging for those exposed to CORT or incubated at lower temperatures.

@@

We manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold -- 23 ± 3ºC or Hot -- 30 ± 3ºC) in a 2x2 factorial design (@fig-Methods A). Eggs were allocated to one of two different developmental stress treatments: CORT-treated eggs were topically supplied with 5µL of CORT dissolved in 100% Ethanol at a final concentration of 10 pg CORT/mL. In the Control treatment eggs received an equal volume of the 100% Ethanol vehicle. We selected those doses based on previous studies where CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs in _Lampropholis delicata_ [@crino2024eggs]. Then, eggs were incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limit of the natural incubation temperatures for this species [@qualls2000post]. The number of eggs per clutch assigned to each hormone and temperature treatment was counterbalanced per clutch.

@@

Two weeks before starting the numerical discrimination task (see below), lizards were moved to the experimental arenas for acclimatisation. The arenas were individual medium-size (41 L x 29.7 W x 22 H cm) plastic containers provided with a shelter (9 L x 6 W x 1.5 H cm) on one of the sides and a water dish in the center (@fig-Methods B). These arenas were placed in 7 different racks located in two experimental rooms. Each rack was associated with its own CCTV system (device model DVR-HP210475), allowing us to record lizards' behaviour during the experiment. Despite the conditions in the experimental rooms were the same as in the colony room, we counterbalanced the number of lizards per treatment in each rack to control for any potential effects of the room or the position of the lizard in the rack. During acclimatisation and throughout the experiment, lizards were fed with only one cricket daily, dusted with calcium and multivitamins (see protocol below), and water was supplied *ad libitum*. The temperature of both rooms was set to 22-24ºC, but we also provided a temperature gradient (24-32°C) using a heat cord along the center of the arena and heat lamps in a 12 h light: 12 h dark cycle.

The numerical discrimination test consisted of a series of two-choice trials where lizards were presented with different numbers of food items. Each choice was placed on one of the two lateral platforms of a white 3D printed device (see @fig-Methods B, C) and then covered with a transparent Petri dish to prevent the lizard from obtaining the food. We performed five tests that differed in the ratio between both choices: 1 cricket VS 4 crickets, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4. The food consisted of small frozen crickets (*A. domestica*) that were thawed and dusted with calcium and multivitamins. We decided to use frozen crickets to avoid movement that would affect lizards' behaviour [@cooper2024tell].

Lizards were fed frozen crickets for one month prior to the experiments to habituate them to eating frozen prey. To habituate lizards to the experimental conditions, frozen crickets were placed on top of one of the platforms and Petri dishes during each feeding for two weeks prior to experimental trials. The side where the food was placed was changed every day to avoid any side bias during the trials. However, to further control for that, the side where we placed the higher number of crickets was randomised between trials and individuals during the tests. We also counterbalanced the number of individuals per treatment tested in the same type of trial each day. The crickets used during the experiment were all the same size, but to control for other quantitative traits [@agrillo2014spontaneous], we changed the orientation of the crickets as shown in @fig-Methods D. In addition, after the experiments, we conducted another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see Supplementary Material - Prey orientation test).

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Experimental design of the study. Panel (**A**) shows the early environment manipulation procedures. (**B**) illustrates the arena where the tests were performed, while panel (**C**) indicates the measurements of the platform used for the experiments. Finally, panel (**D**) displays the types of numerical tests used and the orientation of the crickets in each test."

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) Latency, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (Choice), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) Interest shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of what was the first choice. We measured the interest shown for the higher amount of food (Interest) by substracting the time interacting with the option with fewer crickets to the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as 0 for that option. When there was no interaction with any of the choices, both Latency, Choice, and Interest were noted as NA. We also considered Interest as NA in a few cases where the animals managed to lift the dish and eat the prey; Latency and Choice were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests.

@@

We modelled the three main recorded behaviours: Latency, Choice, and Interest as the response variables. We assumed normal distributions for Latency (log-transformed) and Interest [family = gaussian()]. The error structure for Choice was modeled using a Bernoulli distribution with a logit link function [family = Bernoulli(link = ‘logit’)].  We included test (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4), hormone (CORT versus Control), incubation temperature (Cold versus Hot), and the three-way interaction as fixed factors. We also included in our predictors sex, and age (centered to 0). We included lizard identity and clutch as random factors. _L. guichenoti_ lays up to two clutches per year [@joss1985reproductive]. Since egg collection was done during half of the breeding season, each clutch likely came from a unique mother, and so, clutch identity captures potential maternal effects.

We used the posterior distributions of parameters from these models to test for between treatment differences in the ability to discriminate different numbers of crickets for each of the numerical tests. We used p~mcmc~ to tests whether the contrast between treatments for each test differed from zero and whether Choice and Interest differed from 0.5 or zero respectively. We considered statistical significance if p~mcmc~ < 0.05.

@@

```{r, posteriors}
#| label: posteriors
# Code to extract posteriors from the model and get predicted values
source(here("R", "func.R"))
#
## Extract posteriors
posteriors <- as_draws_df(model)
# 
## Get predictions for every test and treatment based on the posteriors of the model. We used the function org_post (see func.R) to get the predictions.
#
# Latency
post_lat <- org_post("lat")
# Choice
post_choice <- org_post("choice")
prob_choice <- exp(post_choice)/(1 + exp(post_choice))
# Estimated interest for the higher amount of food
post_int <- org_post("int")
#
```

Latency was affected by sex (mean estimated effect of sex = `r {format_dec(mean(posteriors$b_loglatency_sexm), 2)}`, 95% CI = [`r {format_dec(quantile(posteriors$b_loglatency_sexm, 0.025), 2)}`, `r {format_dec(quantile(posteriors$b_loglatency_sexm, 0.975), 2)}`]) and age (mean estimated effect of age = `r {format_dec(mean(posteriors$b_loglatency_age),2)}`, 95% CI = [`r {format_dec(quantile(posteriors$b_loglatency_age, 0.025), 2)}`, `r {format_dec(quantile(posteriors$b_loglatency_age, 0.975),2)}`]; see Table S1 in Supplementary Material). Males and older individuals made, overall, faster decisions. However, we did not see such effects on Choice or Interest (see Table S1 in Supplementary Material). Nevertheless, when we analyse the effect of treatment on performance, for every variable recorded, we control for sex by averaging the predicted values for males and females, and we control for age by including the centered age in the model. These results are summarized in @fig-results and Table 1.

>>>>

```{r, packages}
#| label: packages
pacman::p_load(tidyverse, flextable, emmeans, DHARMa, brms, here, ggplot2, lme4, zoo, lmerTest, broom, tidybayes, forcats, MASS, nortest, fitdistrplus, ggh4x, PupillometryR, cowplot, png, grid, remotes, ggthemes, bayestestR, HDInterval)
```

@@

## Abstract
From social behaviour to navigating their environment, quantitative abilities can be crucial to fitness. Early life conditions can impact brain development affecting quantitative competence. In anthropogenically modified environments, early thermal conditions or glucocorticoid concentrations (GCs) – stress-related hormones transmitted by the parents – can play a prominent role in shaping cognition through their effects on brain physiology. Furthermore, temperature can lead to increased levels of GCs that are passed on to offspring, emphasizing the need to study the combined effects of these two factors. Here, we investigated the effects of elevated prenatal corticosterone (CORT) – the main GC in reptiles – and incubation temperature on numerical discrimination in the common garden skink (_Lampropholis guichenoti_). We then assessed numerical abilities through a spontaneous choice test with food as the relevant stimulus. Employing a repeated measures design, we subjected lizards to five numerical tests each differing in the ratios between the two choices (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 2 VS 4). Contrary to our expectations, we found no evidence of numerical discrimination or any effect of treatments on lizard behaviour. These findings suggest that _L. guichenoti_ may not use numerical cues to make foraging decisions. Our findings align with previous studies indicating that reptiles feeding on live prey may have difficulty distinguishing between different prey quantities in spontaneous choice tests.

## Significance statement
Discriminating between quantities can be crucial for individual fitness, yet the impact of early environmental factors on the development of quantitative competence remains largely unexplored. In this study, we investigated, for the first time, the effects of prenatal corticosterone exposure and incubation temperature on numerical discrimination in a reptile. Contrary to our predictions, we did not observe any evidence of numerical discrimination and therefore no impact of early developmental environment. Our results align with previous studies showing that reptiles consuming live prey struggle to differentiate prey numbers in spontaneous choice tests. These results open new avenues for exploring the evolution of numerical abilities in reptiles and identifying factors influencing the development of quantitative skills in this group. Further research should explore alternative methods, such as trained tasks, to better understand the cognitive mechanisms underlying quantitative competence in reptiles.

## Introduction  
The ability to discern between quantities is a fundamental cognitive skill widely observed across the animal kingdom [@stancher2015numerical; @vonk2012bears; @beran2016capuchin]. Numerical cognition can contribute to social decisions, mate selection, and foraging and thus is adaptive in various ecological contexts [@lucon2017individual; @nieder2018evolution]. For instance, female lions (_Panthera leo_) use numerical information to assess the risk of confronting rival groups [@mccomb1994roaring], male adult mealworms (_Tenebrio molitor_) select sites with scents of more females [@carazo2009quantity], and red-backed salamanders (_Plethodon cinereus_) prefer tubes containing larger numbers of prey [@uller2003salamanders]. Given the adaptive advantages of numerical and quantity discrimination, research has extensively has focused on exploring the ability of different species to differentiate between quantities under different contexts and paradigms.

Numerical competence is often evaluated by considering the types of cues animals use [@plotnik2019elephants; @cooper2024tell], the ecological contexts in which these abilities are deployed [@mccomb1994roaring; @chittka1995can; @uller2003salamanders; @carazo2009quantity], and the neural mechanisms that underlie them [@nieder2005counting]. The latter have demonstrated that non-verbal numerical representations in mammals activates an extensive neural pathway including regions involved in eye movement, attention control, or complex learning [@nieder2005counting]. This connection between numerosity and brain physiology underscores the importance of understanding how brain development influences numerical abilities.

Brains are particularly sensitive to environmental inputs during early stages of development [@coomber_independent_1997; @zhu_prenatal_2004], with long-lasting effects on cognitive abilities [@zhu_prenatal_2004; @amiel_egg_2014]. For instance, glucocorticoids (GCs), steroid hormones that mediate the physiological stress response [@sapolsky_how_2000], can alter animals' performance in spatial, associative, or reversal learning tasks if elevated during early stages of development [@szuran_water_1994; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @lui2017chronic]. Prenatal exposure to GCs is typically associated with impaired learning skills, but these effects have been shown to vary depending on factors such as dose, sex, or type of cognitive test employed [@szuran_water_1994; @farrell_developmental_2015-learn; @bebus_associative_2016; @haq_prenatal_2021]. Nevertheless, little is known about how exposure to GCs during development can affect quantitative abilities.

The effects of prenatal GCs on learning may be modulated by other developmental effects or environmental conditions. For example, the early thermal environment plays a prominent role in brain development and cognitive abilities in ectotherms [@coomber_independent_1997;@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_incubation_2018; @vila_pouca_quantity_2019]. In lizards, incubating eggs at elevated temperatures improves cognitive abilities consistent with elevated neuronal density in some areas of the brain [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017], but impairs cognition at temperatures above lizards’ natural incubation thermal range [@dayananda_incubation_2017; @abayarathna_effects_2020]. In addition, when @vila_pouca_quantity_2019 tested the effects of elevated incubation temperature on numerical discrimination in  Port-Jackson sharks (_Heterodontus portusjacksoni_), they found that sharks incubated at elevated temperatures learned faster in a quantity discrimination task [@vila_pouca_quantity_2019]. However, the effects of early thermal environment on numerical abilities in other taxa remains largely unexplored.

Temperature can affect GC levels in vertebrates and can indirectly affect developing offspring through maternal effects [@jessop2016multiscale; @racic_effects_2020; @mentesana2022glucocorticoids; @Crino_2023]. Early-life GCs and thermal conditions could have interactive effects that influence cognitive development. Understanding how these factors influence quantitative abilities could reveal important insights into the ecological relevance of numerosity. To fill this gap, we used the common garden skink (_Lampropholis guichenoti_) as a model to test the effects of elevated prenatal corticosterone (CORT) - the main GC in reptiles [@Crino_2023] - and incubation temperature on numerical discrimination abilities. However, the combined influence of prenatal CORT and thermal environment on cognition remains unknown, particularly on numerical abilities.

Given the influence of numerical cognition on fitness, we would expect evolutionary pressures to be similar across taxa, and yet, research on numerical discrimination has been traditionally focused on mammals, birds, and fishes [@nieder2005counting; @agrillo2014spontaneous]. However, recent studies demonstrate that reptiles are also competent in numbers. For example, red-footed turtles (_Chelonoidis carbonaria_) rapidly learned to associate colour with higher volumes of food [@soldati2017long], Hermann’s tortoises (_Testudo hermanni_) discriminate between large and small food quantities in a spontaneous choice test [@gazzola2018continuous], and freshwater turtles (_Trachemys scripta_ and _Mauremys sinensis_) rapidly learned a quantity discrimination task [@lin2021superior; @lin2024trained]. The italian wall lizard (_Podarcis siculus_) failed to discriminate between different numbers of food items in a spontaneous choice test [@miletto2017quantitative], but succeeded in a trained quantity discrimination task [@miletto2018quantity]. In contrast, Iberian rock lizards (_Iberolacerta cyreni_) preferred larger quantities of crickets when presented five versus ten items [@recio_prey_2021]. Indeed, the literature shows evidence for basic quantitative abilities in reptiles that may vary with the type of cue and the test employed.

Here, we investigated the effect of prenatal CORT and incubation temeprature on numerical discrimination in a lizard. We manipulated CORT concentration (CORT-treated or a sham control) in _L. guichenoti_ eggs and incubated them under one of two temperature regimes (Cold - 23 ± 3ºC or Hot - 28 ± 3ºC) in a 2x2 factorial design. Then, we assessed numerical discrimination abilities of lizards unsing a spontaneous choice test approach. The numerical discrimination task consisted of five tests that differed in the relative number of food items: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. We based this experiment on observations in nature that the ability to detect a change in stimulus intensity depends on the relative, not absolute, difference between the stimuli ['Weber's law'; @agrillo2014spontaneous]. Built upon other studies that explored the cognitive effects of prenatal GCs or thermal early environment in reptiles and other taxa, we hypothesized that lizards incubated at higher temperatures would outperform those incubated at lower temperatures in the numerical discrimination task [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @vila_pouca_quantity_2019]. We predicted that increased CORT levels would impair numerical competence [@szuran_water_1994; @zhu_prenatal_2004], with more pronounced effects at lower temperatures. Consistent with Weber's law, we predicted that lizards incubated at higher temperatures would successfully discriminate between quantities even at high ratios (e.g. 3 VS 4), whereas even the lowest ratios (e.g. 1 VS 4) could be challenging for those exposed to CORT or incubated at lower temperatures.

@@

We manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (Cold -- 23 ± 3ºC or Hot -- 30 ± 3ºC) in a 2x2 factorial design (@fig-Methods A). Eggs were allocated to one of two different developmental stress treatments: CORT-treated eggs were topically supplied with 5µL of a CORT solution (10pg/mL) Control treatment eggs received an equal volume of 100% Ethanol. Corticosterone treatments were made by dissolving crystalline corticosterone (Sigma, Cat. No. C2505) in 100% ethanol. We selected those doses based on previous studies where CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs in _Lampropholis delicata_ [@crino2024eggs]. Then, eggs were incubated in one of the two previously mentioned temperature regimes (Cold or Hot) until hatching. These temperatures represent the upper and lower limit of the natural incubation temperatures for this species [@qualls2000post]. The number of eggs per clutch assigned to each hormone and temperature treatment was counterbalanced per clutch.

@@

Two weeks before starting the numerical discrimination task (see below), lizards were moved to the experimental arenas for acclimatisation. The arenas were individual medium-size (41 L x 29.7 W x 22 H cm) plastic containers provided with a shelter (9 L x 6 W x 1.5 H cm) on one of the sides and a water dish in the center (@fig-Methods B). Arenas were placed on seven different racks located in two experimental rooms. Each rack was associated with its own CCTV system (device model DVR-HP210475), allowing us to record lizards' behaviour during the experiment. Despite the conditions in the experimental rooms were the same as in the colony room, we counterbalanced the number of lizards per treatment in each rack to control for any potential effects of the room or the position of the lizard in the rack. During acclimatisation and throughout the experiment, lizards were fed with only one cricket daily, dusted with calcium and multivitamins (see protocol below), and water was supplied *ad libitum*. The temperature of both rooms was set to 22-24ºC, but we also provided a temperature gradient (24-32°C) using a heat cord along the center of the arena and heat lamps in a 12 h light: 12 h dark cycle.

The numerical discrimination test consisted of a series of two-choice trials where lizards were presented with different numbers of food items. Each choice was placed on one of the two lateral platforms of a white 3D printed device (see @fig-Methods B, C) and then covered with a transparent Petri dish to prevent the lizard from obtaining the food. We performed five tests that differed in the ratio between both choices of crickets. Specifically, we compared: 1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4 crickets. The food consisted of small frozen crickets (*A. domestica*) that were thawed and dusted with calcium and multivitamins. We decided to use frozen crickets to avoid movement that would affect lizard behaviour [@cooper2024tell].

Lizards were fed frozen crickets for one month prior to the experiments to habituate them to eating frozen prey. To habituate lizards to the experimental conditions, frozen crickets were placed on top of one of the platforms and Petri dishes during each feeding for two weeks prior to experimental trials. The side where the food was placed was changed every day to avoid any side bias during the trials. However, to further control for side biases, the higher number of crickets was randomly placed on the right or the left platform. We also counterbalanced the number of individuals per treatment tested in the same type of trial each day.

The crickets used during the experiment were all the same size, but to control for other quantitative traits we changed the orientation of the crickets as illustrated in @fig-Methods D. This approach created a conflict of information between choices, as horizontally oriented crickets could occupy the same space as a greater number of vertically oriented crickets. This allowed us to determine whether the lizards relied on discrete numerical units rather than non-numerical cues, such as total length, area, or volume [@agrillo2014spontaneous]. In addition, after the experiments, we conducted another set of trials to test whether the orientation of the crickets could affect lizards’ choice, but we found no effect of orientation (see _Supplementary Material - Prey orientation test_).

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Experimental design. Panel (**A**) shows the early environment manipulation procedures. (**B**) illustrates the arena where the tests were performed, while panel (**C**) indicates the measurements of the platform used for the experiments. Finally, panel (**D**) displays the types of numerical tests used and the orientation of the crickets in each test."

@@

All tests were done between 1000-1200 hours when the lizards were most active. Each trial was performed daily for five days between the 17^th^ and 22^nd^ of May 2024. We recorded three main variables: i) _Latency_, defined as the time between the start of the test - when the platform was left in the enclosure - until the first interaction with one of the options; ii) First choice of the lizard (_Choice_), recorded as 1 if the lizard interacted first with the larger amount or 0 if not; and iii) _Interest_ shown for the higher amount of food (detailed below). We considered that the lizard interacted with the food when it touched the Petri dish with its snout or when the front half of the body was on top of the platform. For five minutes after the first interaction with one of the options, we recorded the time spent interacting with that option for a maximum of five minutes; we did this for both amounts of food independently of what was the first choice. We measured the interest shown for the higher amount of food (_Interest_) by subtracting the time interacting with the option with fewer crickets to the time spent interacting with the larger amount of food. If there was no interaction with one of the options during the test, we considered the time spent as zero for that option. When there was no interaction with any of the choices, both _Latency_, _Choice_, and _Interest_ were noted as NA. We also considered Interest as NA in a few cases where the animals managed to lift the dish and eat the prey; Latency and Choice were recorded and employed in the analyses. We excluded from the analyses all the individuals that did not make any choice in three or more tests.

Motivation plays a key role in animals' performance in spontaneous choice tests [@agrillo2014spontaneous; @bisazza2014experim]. To ensure that lizards remained motivated to forage, they were provided with only one cricket during both acclimatization and the numerical task (see above). During acclimatization, lizards were fed at the same time each day as the experiments were conducted. In the numerical task, lizards received a cricket immediately after the test and then fasted for more than 20 hours, which we expected would help standardize fasting levels across individuals. To further control for motivation, we excluded from our analyses any trials in which lizards did not interact with any options or did not eat the cricket provided after the test (n = 0 in both cases). Consequently, we do not consider motivation to be a major factor influencing our results (but see Discussion).

@@

We modelled the three main recorded behaviours: _Latency_, _Choice_, and _Interest_ as the response variables. We assumed normal distributions for Latency (log-transformed) and Interest [family = gaussian()]. The error structure for Choice was modeled using a Bernoulli distribution with a logit link function [family = Bernoulli(link = ‘logit’)]. We included test (1 VS 4, 1 VS 3, 2 VS 4, 2 VS 3, 3 VS 4), hormone (CORT versus Control), incubation temperature (Cold versus Hot), and the three-way interaction as fixed factors. We also included in our predictors sex, and age. We did not find any effect of sex or age on Choice or Interest, but we found an effect on Latency (see Results). However, we decided to include both factors in all models. Age was centered to zero in the model, and we averaged the predicted values for males and females.

We included lizard identity and clutch as random factors. _L. guichenoti_ lays up to two clutches per year [@joss1985reproductive]. Since egg collection was done during half of the breeding season, each clutch likely came from a unique mother, and so, clutch identity captures potential maternal effects.

We used the posterior distributions of parameters from these models to test for between treatment differences in the ability to discriminate different numbers of crickets for each of the numerical tests. Because the posteriors did not follow a normal distribution, we report the median as the main measure of central tendency. We also employed Highest Density Intervals using the hdi function in bayestestR [@bayestestR] to test if the contrasts between treatments for all the variables, or the _Interest_ per each test was different form zero. Given Choice was Bernoulli, we modified our null hypotheses to test if differed from a probability of 0.5 as there were only two possible options.

@@

```{r, posteriors}
#| label: posteriors
# Code to extract posteriors from the model and get predicted values
source(here("R", "func.R"))
#
## Extract posteriors
posteriors <- as_draws_df(model)
# 
## Get predictions for every test and treatment based on the posteriors of the model. We used the function org_post (see func.R) to get the predictions.
#
# Latency
post_lat <- org_post("lat")
normality_results_lat <- check_normality(post_lat)
# Choice
post_choice <- org_post("choice")
prob_choice <- exp(post_choice)/(1 + exp(post_choice))
normality_results_choice <- check_normality(prob_choice)
write.csv(normality_results_choice, file = here("output/Checking/normality_results.csv"))
# Estimated interest for the higher amount of food
post_int <- org_post("int")
normality_results_int <- check_normality(post_int)
#
```

Overall, males showed lower latencies than females (median of the estimated effect of sex = `r {format_dec(median(posteriors$b_loglatency_sexm), 2)}`, Higest Density Interval = [`r {format_dec(hdi(posteriors$b_loglatency_sexm, credMass = 0.95)[1], 2)}`, `r {format_dec(hdi(posteriors$b_loglatency_sexm, credMass = 0.95)[2], 2)}`]), and older individuals also made decisions faster (mean estimated effect of age = `r {format_dec(median(posteriors$b_loglatency_age),2)}`, Higest Density Interval = [`r {format_dec(hdi(posteriors$b_loglatency_age, credMass = 0.95)[1], 2)}`, `r {format_dec(hdi(posteriors$b_loglatency_age, credMass = 0.95)[2],2)}`]; see Table S1 in _Supplementary Material_). However, we did not see any effect of age or sex on _Choice_ or _Interest_ (see _Supplementary Material_).

```{r, fig-results}
#| label: fig-results
#| fig.cap: "Estimates of log-latency (Latency), the probability of choosing the larger number of crickets first (Choice), and the estimated interest in the larger number of crickets (Interest) for each of the numerical tests performed across different developmental treatments. The x-axis represents the estimate, and the y-axis is the posterior density of the estimates. The different colours indicate the different treatments. Points and bars represent the median and Higest Density Intervals of the estimates, respectively. Vertical dashed lines in Choice and Interest graphs values 0.5 and 0, respectively. Asterisks indicate values significantly different from 0."
#
source(here("R", "func.R"))
#
#### A) Organise the df used for both parts of the figure for each variable
#
# First the df for violin plots for each variable using df_violin_plot function from func.r
df_viol_lat <- df_violin_plot("lat")
df_viol_choice <- df_violin_plot("choice")
df_viol_int <- df_violin_plot("int")
#
# Second the df for the points and bars plot for each variable using df_bars_plot function from func.r
df_bars_lat <- df_bars_plot("lat")
df_bars_choice <- df_bars_plot("choice")
df_bars_int <- df_bars_plot("int")
#
#### B) Make the plots for each variable using plotting function from func.r
#
fig_lat <- plotting("lat")
fig_choice <- plotting("choice") 
fig_int <- plotting("int")
#
#### C) Combine the plots and make final figure
fig_results <- plot_grid(fig_lat, fig_choice, fig_int, ncol = 1) +
  annotate("text", x = 0.305, y = 0.56, label = "*", size = 7, color = "black") + 
  annotate("text", x = 0.61, y = 0.505, label = "*", size = 7, color = "black")
ggsave("./output/figures/fig_results.png", plot=fig_results, width = 25, height = 25, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/fig_results.png")
```
